Implementation of a lattice Boltzmann method for immiscible multiphase flow simulations using the level set method

### 2015-11-10 Paper/paper.tex
Kapitel 2 sollte nun inhaltlich passen.

\author
{Lorenz Hufnagel, Daniel Zint\\
Chair for System-Simulation, Friedrich Alexander Univerit\"at Erlangen-N\"urnberg\\
91058 Erlangen, Germany % ist das so richtig?
}

@@

Furthermore, we need boundary conditions and initial conditions. For our domain-boundaries and the interior we can take for example
\begin{equation}
	u_i(x,t) = 0 \text{, on } \delta\Omega_i \backslash \Gamma \text{,}
\end{equation}
\begin{equation}
	u_i(x,0) = u_{ini,i}(x) \text{, in } \Omega_i \text{.}
\end{equation}
For the interface $\Gamma$ we have the influence of the surface tension and we assume continuity of the velocities within the fluids. Therefore, we get the following jump conditions:
\begin{equation}
	[u] = 0 \text{,  } [T]n = 2 \sigma \kappa n \text{.}
	\label{jumpconditions}
\end{equation}
Here, T is the stress tensor, $\sigma$ the surface tension, and $\kappa$ the curvature of the interface with respect to its normal $n$. Brackets denote the jump of a quantity at the interface: $[q](x) = \lim_{\epsilon \rightarrow 0} (q(x + \epsilon n) - (q(x - \epsilon n)) \text{ with } x \in \Gamma$.

@@

\subsection{The Lattice Boltzmann Method}
To solve the incompressible Navier-Stokes equations, we use the lattice Boltzmann method. This method describes the evolution of the particle density $f(x,v,t)$ in phase space, with a $(x,v)$ as the phase space variables and $t$ as time. The LBM algorithm discretizes with a regular grid in space and with restricted number of velocities adapted to this grid \cite{LBM1,LBM2,LBM3,LBM4}. We use the D2Q9 model in 2D, which has 9 velocity vectors on a plane grid with unit spacing including one zero velocity (Fig. \ref{lbmc_i}). 
\begin{figure}
	\hfill\includegraphics[width=6cm,natwidth=352,natheight=352]{LBMc_i.png}\hspace*{\fill}
	\caption{The vectors of one cell in LBM}
	\label{lbmc_i}
\end{figure}
The vectors are given by the columns of the matrix
\begin{equation}
	c = 
	\begin{bmatrix}
	0 & 0 & 1 & 1 &  1 &  0 & -1 & -1 & -1 \\
	0 & 1 & 1 & 0 & -1 & -1 & -1 &  0 &  1
	\end{bmatrix}
	\text{.}
\end{equation}
The corresponding particle distributions are denoted by $f_i(x,t) = f(x,c_i,t)$. We can get the density and velocity from the particle distribution functions by calculating:
\begin{equation}
	\rho(x,t) = \sum_{i=0}^{8} f_i(x,t) \text{, } 
	u = \sum_{i=0}^{8} f_i(x,t)c_i \text{.}
\end{equation}
For the collision step we use the Bhatnagar-Gross-Krook (BGK) approximation \cite{BGK}
\begin{equation}
	f_i(x+c_i,t+1) = f_i(x,t) - \frac1\tau(f_i-f_i^{eq}) + G_i \text{.}
\end{equation}
The parameter $\tau$ is the relaxation parameter for the BGK collision operator and controls the kinematic viscosity $\nu = \frac16(2\tau - 1)$. Furthermore, we use the equilibrium distribution
\begin{equation}
	f_i^{eq}(f) = f_i^*(\rho + 3c_i \cdot u + \frac92(c_i \cdot u)^2 - \frac32u^2)
\end{equation}
with the corresponding D2Q9 weight factors
$$
f_i^* = \begin{cases} \frac49 \text{,} & i = 0 \\ \frac19 \text{,} & i = 1,3,5,7 \\ \frac{1}{36} \text{,} &  i = 2,4,6,8 \end{cases} \text{.}
$$

The LBM algorithm consists out of the following steps:
\begin{itemize}
	\item[1. ] Collision step: $f_i^+ = f_i - \frac1\tau(f_i - f_i ^{eq}) + G_i$
	\item[2. ] Propagation step: $f_i(x+c_i,t+1) = f_i^+(x,t)$, for interior nodes
	\item[3. ] Boundary conditions: $f_i(x+c_i,t+1) = f_i(x+c_i,t)$, if $x \notin \Omega$
\end{itemize}

@@

To solve this equation we use the Toolbox of Ian Mitchell \cite{mitchell}. This toolbox takes as input the velocity field we get from the lattice Boltzmann method and computes the movement of the interface. The toolbox also provides functions to calculate the normal and the curvature of the surface. We will need this informations for the coupling of the methods.

\subsection{Coupling of LBM and level set method}
The lattice Boltzmann method requires a boundary treatment at the interface which implements the jump conditions (\ref{jumpconditions}). We use a new boundary condition that is based on a simple bounce back type Dirichlet condition. For example, in Fig. \ref{intro_phases} for a point $x_2 \in \Omega_2$ that is next to the boundary we set
% TODO: zweites Bild einfügen mit x_1, x_2 und c_i

@@

	S_k = -\frac{3}{2\tau h^2} \sum_i c_i \otimes c_i(f_i-f_i^{eq})(t,x_k) \text{,}

@@

	\Lambda_i:[S] = ([S]:n \otimes n)((n \cdot c_i)^2 - {|c_i|}^2/3) + 2([S]:n \otimes t_k)(n \cdot c_i)(t_k \cdot c_i) \text{.}

@@

For the calculation we use

@@

Here $n$ is the normal of the interface, $t$ the tangential, $\kappa$ the curvature, and $\mu$ the dynamic viscosity.

@@

	[p] \approx \frac{1}{3 h^2} (\rho(x_1,t)\varrho_1 - \rho(x_2,t)\varrho_2) \text{.}

@@

We use now the information of the level set method for the new boundary condition. What is still missing, is a refill method. Due to the fact that the interface moves, some cells change their fluid type. We use an equilibrium/non-equilibrium refill as it is described in \cite{Caiazzo} that consists out of four steps:

@@

Suppose we choose an inward pointing direction $c_i$ in a boundary point $x = x_b \in \Omega_1$, e.g. the direction of biggest angle with the surface normal. Then the density is interpolated by

@@

\begin{figure}[h]

@@

We can now calculate the new equilibrium $f_i^{eq}(\tilde{\rho},\tilde{u})$ and add it to the non-equilibrium part that we copy from the direct neighbor, $f_i^{neq}(x+c_i,t+1)$.
The same procedure has also to be applied in the second fluid $\Omega_2$.
\begin{figure}[h]

@@

Finally we couple the two methods in the following way:
	\item[1.] Create the initial interface $\Gamma_0$
	\item[2.] Run the level set code once to create the surface description for LBM
	\item[3.] Run the LBM using the data from the level set for a prescribed number of steps
	\item[4.] Pass the current velocity field to the level set code and run it for the same number of steps (see Fig. \ref{dataexchange})
	\item[5.] Repeat step 3 and 4 until the end of the simulation.

>>>>

\author
{Lorenz Hufnagel, Daniel Zint\\
Chair for System-Simulation, Friedrich Alexander Univerit\"at Erlangen-N\"urnberg\\
Cauerstr. 11, 91058 Erlangen, Germany\\
lorenz.hufnagel@studium.uni-erlangen.de\\
daniel.zint@fau.de % ist das so richtig?
}

@@

Furthermore, we need boundary conditions and initial conditions. For our domain-boundaries and the interior we can take for example
\begin{equation}
	u_i(x,t) = 0 \text{, on } \delta\Omega_i \backslash \Gamma \text{,}
\end{equation}
\begin{equation}
	u_i(x,0) = u_{ini,i}(x) \text{, in } \Omega_i \text{.}
\end{equation}
For the interface $\Gamma$ we have the influence of the surface tension and we assume continuity of the velocities within the fluids. Therefore, we get the following jump conditions:
\begin{equation}
	[u] = 0 \text{,  } [T]n = 2 \sigma \kappa n \text{.}
	\label{jumpconditions}
\end{equation}
Here, T is the stress tensor, $\sigma$ the surface tension, and $\kappa$ the curvature of the interface with respect to its normal $n$. Brackets denote the jump of a quantity at the interface: 
$[q](x) = \lim_{\epsilon \rightarrow 0} (q(x - \epsilon n) - (q(x + \epsilon n)) \text{ with } x \in \Gamma$ % I also turned this around, according to [p].
. In Fig. \ref{intro_phases} this would be the jump from $\Omega_1$ to $\Omega_2$. For the other direction the sign changes.

@@

\subsection{The Lattice Boltzmann Method}
To solve the incompressible Navier-Stokes equations, we use the lattice Boltzmann method. This method describes the evolution of the particle density $f(\vec{x},\vec{v},t)$ in phase space, with a $(\vec{x},\vec{v})$ as the phase space variables and $t$ as time. The LBM algorithm discretizes with a regular grid in space and with restricted number of velocities adapted to this grid \cite{LBM1,LBM2,LBM3,LBM4}. We use the D2Q9 model in 2D, which has 9 velocity vectors on a plane grid with unit spacing including one zero velocity (Fig. \ref{lbmc_i}). 
\begin{figure}
	\hfill\includegraphics[width=6cm,natwidth=403,natheight=409]{LBMc_i.png}\hspace*{\fill}
	\caption{The vectors of one cell in LBM}
	\label{lbmc_i}
\end{figure}
The vectors are given by the columns of the matrix
\begin{equation}
	c = 
	\begin{bmatrix}
	0 & 0 & 1 & 1 &  1 &  0 & -1 & -1 & -1 \\
	0 & 1 & 1 & 0 & -1 & -1 & -1 &  0 &  1
	\end{bmatrix}
	\text{.}
\end{equation}
The corresponding particle distributions are denoted by $f_i(\vec{x},t) = f(\vec{x},\vec{c_i},t)$. We get the density and velocity from the particle distribution functions with:
\begin{equation}
	\rho(\vec{x},t) = \sum_{i=0}^{8} f_i(\vec{x},t) \text{, } 
	\vec{u} = \sum_{i=0}^{8} f_i(\vec{x},t)\vec{c_i} \text{.}
\end{equation}
For the collision step we use the Bhatnagar-Gross-Krook (BGK) approximation \cite{BGK}
\begin{equation}
	f_i(\vec{x}+\vec{c_i},t+1) = f_i(\vec{x},t) - \frac1\tau(f_i-f_i^{eq}) + G_i \text{.}
\end{equation}
The parameter $\tau$ is the relaxation parameter for the BGK collision operator and controls the kinematic viscosity $\nu = \frac16(2\tau - 1)$. $G_i$ is an additional force like gravity or a magnetic field. In our case we set $G_i = 0$. Furthermore, we use the equilibrium distribution
\begin{equation}
	f_i^{eq}(\rho,\vec{u}) = f_i^*(\rho + 3\vec{c_i} \cdot \vec{u} + \frac92(\vec{c_i} \cdot \vec{u})^2 - \frac32{|\vec{u}|}^2)
\end{equation}
with the corresponding D2Q9 weight factors
$$
f_i^* = \begin{cases} \frac49 \text{,} & i = 0 \\ \frac19 \text{,} & i = 1,3,5,7 \\ \frac{1}{36} \text{,} &  i = 2,4,6,8 \end{cases} \text{.}
$$

The LBM algorithm consists out of the following steps:
\begin{itemize}
	\item[1. ] Collision step: $f_i^+ = f_i - \frac1\tau(f_i - f_i ^{eq}) + G_i$
	\item[2. ] Propagation step: $f_i(\vec{x}+\vec{c_i},t+1) = f_i^+(x,t)$, for interior nodes
	\item[3. ] Boundary conditions: $f_i(\vec{x}+\vec{c_i},t+1) = f_i(\vec{x}+\vec{c_i},t)$, if $\vec{x} \notin \Omega$
\end{itemize}

@@

To solve this equation we use the Toolbox of Ian Mitchell \cite{mitchell} for Matlab. This toolbox takes as input the velocity field we get from the lattice Boltzmann method and computes the movement of the interface. The toolbox also provides functions to calculate the normal and the curvature of the surface. We will need this informations for the coupling of the methods.

\subsection{Coupling of LBM and level set method}
The lattice Boltzmann method requires a boundary treatment at the interface which implements the jump conditions (\ref{jumpconditions}). For sake of simplicity, we don't write arrows for every vector, i.e. $\vec{u} \rightarrow u$. One should consider that every position, direction, and velocity is two-dimensional.

For the coupling, we create a new boundary condition that is based on a simple bounce back type Dirichlet condition. For example, in Fig. \ref{coupling} for a point $x_2 \in \Omega_2$ that is next to the boundary we set
\begin{figure}
	\hfill\includegraphics[width=10cm,natwidth=437,natheight=178]{coupling.png}\hspace*{\fill}
	\caption{link between two points in different fluids}
	\label{coupling}

@@

\Lambda_i = c_i \otimes c_i - \frac13 {|c_i|}^2 \mathbf{I} \text{,}

@@

	S_k = -\frac{3}{2\tau h^2} \sum_i c_i \otimes c_i(f_i-f_i^{eq})(t,x_k) \text{.}

@@

The tensor product is $a \otimes b = ab^T$ and the double contraction $A:B = \mathrm{trace}(AB^T)$. Furthermore, we have
	\Lambda_i:[S] = ([S]:n \otimes n)((n \cdot c_i)^2 - {|c_i|}^2/3) + 2([S]:n \otimes t_k)(n \cdot c_i)(t_k \cdot c_i) \text{,}
with the terms

@@

Here $n$ is the normal of the interface, $t$ the tangential, $\bar{S} = (S_1 + S_2)/2$ the averaged symmetric velocity gradient, $\kappa$ the curvature, and $\mu$ the dynamic viscosity, with $[\mu] = \mu_2 - \mu_1$ as its jump.

@@

	[p] \approx \frac{1}{3 h^2} (\rho(x_2,t)\varrho_2 - \rho(x_1,t)\varrho_1) \text{.}

@@

What is still missing, is a refill method. Due to the fact that the interface moves, some cells change their fluid type. We use an equilibrium/non-equilibrium refill as it is described in \cite{Caiazzo} that consists out of four steps:

@@

Suppose we choose an inward pointing direction $c_i$ in a boundary point $x \in \Omega_1$, e.g. the direction of biggest angle with the surface normal. Then the density is interpolated by

@@

\begin{figure}

@@

We can now calculate the new equilibrium $f_i^{eq}(\tilde{\rho},\tilde{u})$ and add it to the non-equilibrium part that we copy from the direct neighbor:
$$f_i(\vec{x},t+1) = f_i^{eq}(\tilde{\rho},\tilde{u}) + f_i^{neq}(x+c_i,t+1) \text{.}$$
We have to repeat the procedure for other fluid.
\begin{figure}

@@

Finally, we can couple the two methods in the following way:
\begin{itemize}
	\item[1.] Create the initial interface $\Gamma_0$
	\item[2.] Run the level set code once to create the surface description for LBM
	\item[3.] Run the LBM using the data from the level set for a prescribed number of steps
	\item[4.] Pass the current velocity field to the level set code and run it for the same number of steps (see Fig. \ref{dataexchange})
	\item[5.] Repeat step 3 and 4 until the end of the simulation.
\end{itemize}
